A sound and adequate game semantics for PCF with countable nondeterminism

### 2017-03-10 nondeterminism/paper/nondeterminism.tex
Major changes.

\documentclass{article}

\input{../common/preamble}

\newcommand{\stle}{{\;\le_s\;}}
\newcommand{\exle}{\sqsubseteq}
\newcommand{\exlub}{\bigsqcup}
\newcommand{\dv}{{\text{\lightning}}}
\DeclareMathOperator{\pocl}{pocl}
\newcommand{\opto}{\longrightarrow}
\newcommand{\n}{{\mathtt{n}}}
\newcommand{\IfO}{{\mathtt{If0}\;}}
\newcommand{\suc}{{\mathtt{suc\;}}}
\newcommand{\0}{{\mathtt{0}}}
\newcommand{\plot}{\triangleleft}

\title{Transfinite Games for Countable Nondeterminism}

\begin{document}

@@

\begin{definition}
  Let $A=(M_A,\lambda_A,\zeta_A,P_A)$ be a game.  Then a (nondeterministic) \emph{strategy} for $A$ is a pair $\sigma=(T_\sigma,D_\sigma)$ such that
  \begin{itemize}
    \item $T_\sigma\subset P_A^P$.
    \item $D_\sigma\subset P_A$ is a postfix-closed subset such that all of the minimal positions in $D_\sigma$ are replies to moves in $T_\sigma$ or limits of moves in $T_\sigma$.  In other words, if $s\in D_\sigma$, then there is some $O$-position $t\prefix s$ such that $t\in D_\sigma$ and all $P$-prefixes of $t$ are contained in $T_\sigma$.
  \end{itemize}

@@

\begin{lemma}\label{signProfilesLemma}
  Let $A,B$ be games.  Given a position $s\in P_A\|P_B$, we define the \emph{sign profile} of $s$ to be given by the signs of its two components:
  \[
    \begin{array}{cc|c}
      \zeta_A(s\vert_A) & \zeta_B(s\vert_B) & \textrm{sign profile} \\
      \hline
      P & P & PP \\
      O & P & OP \\
      P & O & PO \\
      O & O & OO
    \end{array}
    \]
  If $s\in P_{A\tensor B}$ is a finite position, then $s$ cannot have sign profile $OO$.  If $s\in P_{A\implies B}$ is a finite position then $s$ cannot have sign profile $OP$.  
  \begin{proof}
    By negativity, the empty sequence $\epsilon$ always has sign profile $PP$.  Otherwise, $s=s'a$ for some sequence $s'$ and some move $a$.  If $s\in P_{A\tensor B}$ and $s$ has sign profile $OO$, then $s$ is an $O$-position in $A\tensor B$.  Therefore, $s'$ is a $P$-position in $A\tensor B$ and so $s'$ must have sign profile $PP$, by the definition of $\zeta_{A\tensor B}$.  But this is impossible, as the move $a$ can only change the sign profile in one of the two components.

    The proof for the linear implication is entirely similar.
  \end{proof}
\end{lemma}

\begin{proof}[Proof of Proposition \ref{TensorImpliesWellFormed}]
  Alternation is part of the definition and negativity is implied by the equations $P\wedge P = P$ and $P\Rightarrow P = P$.  The only tricky part is proving consistency of the $\lambda$ and $\zeta$ functions.  Since consistency applies only to positions with a last move, we may ignore the infinite plays in $A$ and $B$.

@@

  The proof for the linear implication is similar.
\end{proof}

@@

\begin{proposition}
  Let $A,B,C$ be games, let $\sigma$ be a strategy for $A\implies B$ and let $\tau$ be a strategy for $B\implies C$.  Suppose $s\in\comp\tau\sigma$ is finite.  Then $s$ has even length.
  \begin{proof}
    Write $s=\s\vert_{A,C}$, where $\s\in\sigma\|\tau$.  Then we have
    \begin{align*}
      \length(s) & = \length(\s) - \length(\s\vert_B) \\
       & = \length(\s\vert_{A,B}) + \length(\s\vert_{B,C} - 2\length(\s\vert_B)
    \end{align*}
    $\s\vert_{A,B}$ and $\s\vert_{B,C}$ both have even length, since they are $P$-positions, so the result follows.
  \end{proof}
\end{proposition}

@@

  \begin{proof}
    As in \cite{abramskyjagadeesangames}.  Define
    \[
      \sigma\|\tau\|\upsilon = \left\{\s\in\seqs{(M_A \cprd M_B \cprd M_C \cprd M_D)} \suchthat \s\vert_{A,B}\in\sigma,\;\s\vert_{B,C}\in\tau,\;\s\vert_{C,D}\in\upsilon\right\}
      \]
    We shall show that 
    \[
      T_{\comp{\upsilon}{(\comp{\tau}{\sigma})}} = \{\s\vert_{A,D}\suchthat\s\in\sigma\|\tau\|\upsilon\}
      \]
    The same claim will follow for $T_{\comp{(\comp{\upsilon}{\tau})}{\sigma}}$ by symmetry, completing the proof.

@@

    In any case, we have
    \begin{gather*}
      \u\vert_{A,C,D} = \s_0D_0\s_1D_1\dots = \s \\
      \u\vert_{A,B,C} = \s_0B_0\s_1B_1\dots = \t
    \end{gather*}
    It follows that $\u\in\sigma\|\tau\|\upsilon$: 
    \begin{gather*}
      \u\vert_{A,B} = \t\vert_{A,B} \in \sigma \\
      \u\vert_{B,C} = \t\vert_{B,C} \in \tau \\
      \u\vert_{C,D} = \s\vert_{C,D} \in \upsilon
    \end{gather*}
    Lastly, we have $\u\vert_{A,D} = \s\vert_{A,D} = s$, completing the proof.
  \end{proof}
\end{proposition}

@@

\begin{proposition}
  $D_{\comp\tau\sigma}$ is a well-defined divergence set.
  \begin{proof}
    $D_{\comp\tau\sigma}$ alternating by the same argument we used for $T_{\comp\tau\sigma}$, and is postfix-closed in $P_{A\implies C}^O$ by definition.  Now let $s\in D_{\comp\tau\sigma}$ and suppose that $\s\vert_{A,C}\prefix s$, where $\s\in \sigma\dv\tau$.  We need to show that there is some $O$-position $ta\prefix s$ such that $ta\in D_{\comp\tau\sigma}$ and $t\in T_{\comp\tau\sigma}$.

@@

    The case where $\s\vert_{A,B}\in T_\sigma$ and $\s\vert_{B,C}\in D_\tau$ is similar.
  \end{proof}
\end{proposition}

@@

\begin{proposition}
  Let $A,B,C$ be games, let $\sigma$ be a deterministic strategy for $A\implies B$ and let $\tau$ be a deterministic strategy for $B\implies C$.  Then $\comp\tau\sigma$ is a deterministic strategy.  Moreover, for any $s\in\comp\tau\sigma$, there is a unique minimal $\s\in\sigma\|\tau$ such that $\s\vert_{A,C}=s$.
  \begin{proof}
    The idea behind the proof is that determinism of $\sigma$ and $\tau$ mean that the play in the intermediate game $B$ is completely determined (since every move played in $B$ is either a $P$-move in $A\implies B$ or a $P$-move in $B\implies C$.  For full details, see \cite{abramskyjagadeesangames}.
  \end{proof}
\end{proposition}

@@

\begin{proposition}
  Let $A,B,C$ be games and let $\sigma\from A\implies B$ and $\tau\from B\implies C$ be compact deterministic strategies.  Then $\comp\tau\sigma$ is compact.
  \begin{proof}
    Let $s\in M_{A\implies C}^\omega$ be an infinite sequence such that $t\in\comp\tau\sigma$ for all finite prefixes $t\ppprefix s$.  For each finite prefix $t$ of $s$, there is a unique minimal sequence $\t\in\sigma\|\tau$ such that $\t\vert_{A,C}=t$.  Writing $s$ as the limit of its finite $P$-prefixes
    \[
      t_1 \prefix t_2 \prefix t_3 \prefix \dots
      \]
    we build up a sequence
    \[
      \t_1 \prefix \t_2 \prefix \t_3 \prefix \dots
      \]
    where $\t_i$ is the unique minimal sequence in $\sigma\|\tau$ such that $\t_i\vert_{A,C}=t_i$.  We have $\t_i\prefix\t_{i+1}$ because if we remove the last move from $\t_{i+1}$ then the resulting sequence $\u_i$ satisfies $\u_i\vert_{A,C}=t_i$ and therefore $\t_i\prefix\u_i\prefix\t_{i+1}$ by minimality of $\t_i$.

    Then the limit of the sequences $\t_i\vert_{A,B}$ must be contained in $T_\sigma\cup D_\sigma$, since $\sigma$ is compact, and similarly the limit of the sequences $\t_i\vert_{B,C}$ must be contained in $T_\tau\cup D_\tau$, since $\tau$ is compact.  It follows that the limit $\t$ of the sequences $\t_i$ is contained in $\sigma\|\tau\cup\sigma\dv\tau$, and therefore that $\t\vert_{A,C}=s\in T_{\comp\tau\sigma}\cup D_{\comp\tau\sigma}$.
  \end{proof}
\end{proposition}

@@

\begin{proposition}
  Let $A,B,C$ be games and let $\sigma\from A\implies B$ and $\tau\from B\implies C$ be (nondeterministic, locally compact) strategies.  Then $\comp\tau\sigma$ is locally compact.
  \begin{proof}
    Let $s\in T_{\comp\tau\sigma}$.  Then there is some sequence $\s\in\sigma\|\tau$ such that $\s\vert_{A,C}=s$.  We have $\s\vert_{A,B}\in\sigma$ and $\s\vert_{B,C}\in\tau$, so, by local compactness for $\sigma$ and $\tau$, there are compact strategies $\sigma'\from A\implies B$ and $\tau'\from B\implies C$ such that $\s\vert_{A,B}\in\sigma'$ and $\s\vert_{B,C}\in\tau'$.  After removing positions from (and possibly adding divergences to) $\sigma'$ and $\tau'$, we may assume that they are deterministic (and still compact).  Then $\comp{\tau'}{\sigma'}$ is compact and we have $\s\vert_{A,C}\in\comp{\tau'}{\sigma'}\subset\comp\tau\sigma$ and $D_{\comp\tau\sigma}\subset D_{\comp{\tau'}{\sigma'}}$.
  \end{proof}
\end{proposition}

@@

  \begin{proof}
    Suppose that $s\in\comp\tau\sigma$, that $sa\in P_{A\implies C}$ and that $s\not\in D_{\comp\tau\sigma}$.  Let $\s\in\sigma\|\tau$ be such that $\s\vert_{A,C}=s$.  By local compactness of $\sigma$ and $\tau$, there must be compact strategies $\sigma'\from A\implies B$ and $\tau'\from B\implies C$ such that $\s\vert_{A,C}\in\sigma'$ and $\s\vert_{B,C}\in\tau'$.

@@

    Since $sa$ is an $O$-position, one of $\s a\vert_{A,B}$ and $\s a\vert_{B,C}$ must be an $O$-position (by transitivity of $\Rightarrow$).  Suppose without loss of generality that $\s a\vert_{A,B}$ is an $O$-position.  Since $\s a\vert_{A,B}\not\in D_{\sigma'}$, there must be some move $b$ such that $\s a\vert_{A,B}b\in\sigma'$, since $\sigma'$ satisfies the diverge-or-reply rule.  Then we have $\s ab\vert_{A,B}\in\sigma'$.  Moreover, Lemma \ref{signProfilesLemma} tells us that $\s a\vert_{B,C}$ must be a $P$-position, so $\s ab\vert_{B,C}=\s\vert_{B,C}\in\tau'$.  Therefore, $\s ab\in\sigma'\|\tau'$.  By maximality of $\s a$, $b$ must be a move in $A$ and therefore we have $sab\in\comp{\tau'}{\sigma'}\subset\comp\tau\sigma$.
  \end{proof}
\end{proposition}

@@

\begin{proposition}
  Let $A,B,C,D$ be games, let $\sigma$ be a strategy for $A\implies B$, let $\tau$ be a strategy for $B\implies C$ and let $\upsilon$ be a strategy for $C\implies D$.  Then 
  \[
    D_{\comp\upsilon{(\comp\tau\sigma)}} = D_{\comp{(\comp\upsilon\tau)}\sigma}
    \]
  \begin{proof}
    Let
    \[
      D = \left\{\s\in \seqs{(M_A \cprd M_B \cprd M_C \cprd M_D)} \;\middle|\;\mbox{\pbox{\textwidth}{\flushRight
        either \textbf{1} $\s\vert_{A,B}\in D_\sigma$, $\s\vert_{B,C}\in T_\tau$ and $\s\vert_{C,D}\in T_\upsilon$ \\
        or \textbf{2} $\s\vert_{A,B}\in T_\sigma$, $\s\vert_{B,C}\in D_\tau$ and $\s\vert_{C,D}\in T_\upsilon$ \\
        or \textbf{3} $\s\vert_{A,B}\in T_\sigma$, $\s\vert_{B,C}\in T_\tau$ and $\s\vert_{C,D}\in D_\upsilon$
      }}\right\}
      \]
    We claim that $D_{\comp\upsilon{(\comp\tau\sigma)}}=\pocl_{A\implies D}^O\left(\{\s\vert_{A,D}\suchthat\s\in D\}\right)$.  The same claim will hold for $D_{\comp{(\comp\upsilon\tau)}\sigma}$ by symmetry, completing the proof.

@@

    By the argument from Proposition \ref{traceAssociativity}, we may interleave these two sequences $\t$ and $u$ to form a sequence $\s\in\seqs{(M_A\cprd M_B\cprd M_C\cprd M_D)}$ such that $\s\vert_{A,B,C}=\u$ and $\s\vert_{A,C,D}=\t$.  Then it is clear that $\s\in D$ in all cases and we have $\s\vert_{A,D}=\t\vert_{A,D}\prefix s$.
  \end{proof}
\end{proposition}

@@

\begin{theorem}
  There is a category $\G$ where the objects are games as defined above, morphisms from a game $A$ to a game $B$ are given by nondeterministic strategies $(T_\sigma, D_\sigma)$ for $A\implies B$ and the composition and identity morphisms are as defined above.
  \begin{proof}
    All that is left to do is to show that $\id_A$ really is an identity morphism.  We shall show that if $A$ and $B$ are games and $\sigma$ is a strategy for $A\implies B$ then $\comp\sigma{\id_A}=\sigma$; it will follow by a symmetrical argument that $\comp{\id_B}\sigma = \sigma$.

@@

    Conversely, suppose that $s\in D_\sigma$.  We know already that there is some $\s\in\id_A\|\sigma$ such that $\s\vert_{A_1,A_2}\in\id_A$ and $\s\vert_{A_2,B}\in\sigma$.  Then $\s\vert_{A,C}\in D_{\comp\sigma{\id_A}}$ by definition.
  \end{proof}
\end{theorem}

@@

We make the category $\G$ into a symmetric monoidal closed category with tensor product given by $\tensor$ and linear implication given by $\implies$.  At the level of morphisms, the action of $\blank\tensor\blank$ is defined as follows: if $A,B,C,D$ are games, $\sigma$ is a strategy for $A\implies C$ and $\tau$ is a strategy for $B\implies D$, then $\sigma\tensor\tau$ is the strategy for $(A\tensor B)\implies (C\tensor D)$ given by playing the strategy $\sigma$ in $A$ and $C$ and playing the strategy $\tau$ in $B$ and $D$:
\[
  \sigma\tensor\tau = \left\{s\in P_{(A\tensor B)\implies (C\tensor D)}\suchthat s\vert_{A,C}\in\sigma,\;s\vert_{B,C}\in\tau\right\}
  \]
The divergences for $\sigma\tensor\tau$ are defined using the divergences for $\sigma$ and $\tau$:
\[
  D_{\sigma\tensor\tau} = \pocl^O\left(\left\{s\in P_{(A\tensor B)\implies (C\tensor D)}\;\middle|\;\mbox{\pbox{\textwidth}{\flushRight
    either $s\vert_{A,C}\in D_\sigma$ and $s\vert_{B,D}\in T_\tau$ \\
    or $s\vert_{A,C}\in T_\sigma$ and $s\vert_{B,D}\in D_\tau$
    }}\right\}\right)
  \]
\begin{proposition}
  $\sigma\tensor\tau$ is a valid strategy for $(A\tensor B)\implies (C\tensor D)$
  \begin{proof}
    $\sigma\tensor\tau$ is non-empty and prefix-closed, since $\sigma$ and $\tau$ are.

@@

    We can show that
    \[
      D_{\comp{(\sigma\tensor\tau)}{(\sigma'\tensor\tau')}} = D_{(\comp\sigma{\sigma'})\tensor(\comp\tau{\tau'})}
      \]
    using a similar argument.
  \end{proof}
\end{proposition}

@@

\begin{proposition}
  $\oc A$ is a well formed game.
  \begin{proof}
    Exactly the same argument as in Proposition \ref{TensorImpliesWellFormed}.  We need to prove a version of Lemma \ref{signProfilesLemma} which will tell us that if $s\in P_{\oc A}$ then either $\zeta_A(s\vert_n)=O$ for at most one value of $n$.  
  \end{proof}
\end{proposition}

@@

\subsection{Soundness and adequacy}

Our proofs for soundness and adequacy follow \cite{LairdOrdinalGames}.

>>>>

\documentclass{entcs} \usepackage{prentcsmacro}
\sloppy
% The following is enclosed to allow easy detection of differences in
% ascii coding.
% Upper-case    A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
% Lower-case    a b c d e f g h i j k l m n o p q r s t u v w x y z
% Digits        0 1 2 3 4 5 6 7 8 9
% Exclamation   !           Double quote "          Hash (number) #
% Dollar        $           Percent      %          Ampersand     &
% Acute accent  '           Left paren   (          Right paren   )
% Asterisk      *           Plus         +          Comma         ,
% Minus         -           Point        .          Solidus       /
% Colon         :           Semicolon    ;          Less than     <
% Equals        =3D           Greater than >          Question mark ?
% At            @           Left bracket [          Backslash     \
% Right bracket ]           Circumflex   ^          Underscore    _
% Grave accent  `           Left brace   {          Vertical bar  |
% Right brace   }           Tilde        ~

\input{preamble}

\newcommand{\stle}{{\;\le_s\;}}
\newcommand{\steq}{{\;=_s\;}}
\newcommand{\exle}{\sqsubseteq}
\newcommand{\exlub}{\bigsqcup}
\newcommand{\dv}{{\lightning}}
\DeclareMathOperator{\pocl}{pocl}
\newcommand{\opto}{\longrightarrow}
\newcommand{\n}{{\mathtt{n}}}
\newcommand{\IfO}{{\mathtt{If0}\;}}
\newcommand{\suc}{{\mathtt{suc\;}}}
\newcommand{\0}{{\mathtt{0}}}
\newcommand{\plot}{\triangleleft}
\newcommand{\shad}{\mathfrak{S}}
\newcommand{\tree}{\mathfrak{T}}
\newcommand{\Tau}{T}
\newcommand{\Chi}{X}

\def\lastname{Gowers}
\begin{document}
\begin{frontmatter}
  \title{A sound and adequate game semantics for PCF with countable nondeterminism} \author{John
    Gowers\thanksref{ALL}\thanksref{myemail}}
  \address{Department of Computer Science \\ University of Bath \\
    Bath, United Kingdom} \thanks[ALL]{The author is indebted to
    his supervisor, Jim Laird, without whose research and guidance
    this paper could not have been written.} \thanks[myemail]{Email:
    \href{mailto:wjg27@bath.ac.uk} {\texttt{\normalshape
        W.J.Gowers@bath.ac.uk}}} 
\begin{abstract} 
  We give a sound and adequate game semantics for a simple extension of PCF with countable nondeterminism, given in \cite{LairdOrdinalGames}.  The model is based on the model of idealized Algol with finite nondeterminism given by Harmer and McCusker \cite{mccHFiniteND} with extra information about infinite plays in order to accommodate infinite nondeterminism.  Our proof of soundness for the model is standard, but the proof of adequacy makes use of a new technique that is linked to the possibly transfinite complexity of evaluation trees in the presence of countable nondeterminism.
\end{abstract}
\begin{keyword}
  Game semantics, countable nondeterminism, ordinal complexity
\end{keyword}
\end{frontmatter}

@@

\begin{definition}
  Let $A=(M_A,\lambda_A,\zeta_A,P_A)$ be a game.  Then a (nondeterministic) \emph{strategy} for $A$ is a pair $\sigma=(T_\sigma,D_\sigma)$ such that
  \begin{itemize}
    \item $T_\sigma\subset P_A^P$ is prefix-closed.
    \item $D_\sigma\subset P_A$ is a postfix-closed subset such that all of the minimal positions in $D_\sigma$ are replies to moves in $T_\sigma$ or limits of moves in $T_\sigma$.  In other words, if $s\in D_\sigma$, then there is some $O$-position $t\prefix s$ such that $t\in D_\sigma$ and all $P$-prefixes of $t$ are contained in $T_\sigma$.
  \end{itemize}

@@

\begin{lemma}\label{signProfilesLemma}
  Let $A,B$ be games.  Given a position $s\in P_A\|P_B$, we define the \emph{sign profile} of $s$ to be given by the signs of its two components:
  \[
    \begin{array}{cc|c}
      \zeta_A(s\vert_A) & \zeta_B(s\vert_B) & \textrm{sign profile} \\
      \hline
      P & P & PP \\
      O & P & OP \\
      P & O & PO \\
      O & O & OO
    \end{array}
    \]
  If $s\in P_{A\tensor B}$ is a finite position, then $s$ cannot have sign profile $OO$.  If $s\in P_{A\implies B}$ is a finite position then $s$ cannot have sign profile $OP$.  
\end{lemma}
\begin{proof}
  By negativity, the empty sequence $\epsilon$ always has sign profile $PP$.  Otherwise, $s=s'a$ for some sequence $s'$ and some move $a$.  If $s\in P_{A\tensor B}$ and $s$ has sign profile $OO$, then $s$ is an $O$-position in $A\tensor B$.  Therefore, $s'$ is a $P$-position in $A\tensor B$ and so $s'$ must have sign profile $PP$, by the definition of $\zeta_{A\tensor B}$.  But this is impossible, as the move $a$ can only change the sign profile in one of the two components.

  The proof for the linear implication is entirely similar.
\end{proof}

\begin{proof*}{Proof of Proposition \ref{TensorImpliesWellFormed}}
  Alternation is part of the definition and negativity is implied by the equations $P\wedge P = P$ and $P\Rightarrow P = P$.  The only tricky part is proving consistency of the $\lambda$ and $\zeta$ functions.  Since consistency applies only to positions with a last move, we may ignore the infinite plays in $A$ and $B$.

@@

  The proof for the linear implication is similar.
\end{proof*}

@@

\begin{proposition}
  Let $A,B,C$ be games, let $\sigma$ be a strategy for $A\implies B$ and let $\tau$ be a strategy for $B\implies C$.  Suppose $s\in\comp\tau\sigma$ is finite.  Then $s$ has even length.
\end{proposition}
\begin{proof}
  Write $s=\s\vert_{A,C}$, where $\s\in\sigma\|\tau$.  Then we have
  \begin{align*}
    \length(s) & = \length(\s) - \length(\s\vert_B) \\
     & = \length(\s\vert_{A,B}) + \length(\s\vert_{B,C} - 2\length(\s\vert_B)
  \end{align*}
  $\s\vert_{A,B}$ and $\s\vert_{B,C}$ both have even length, since they are $P$-positions, so the result follows.
\end{proof}

@@

\end{proposition}
  \begin{proof}
    As in \cite{abramskyjagadeesangames}.  Define
    \[
      \sigma\|\tau\|\upsilon = \left\{\s\in\seqs{(M_A \cprd M_B \cprd M_C \cprd M_D)} \suchthat \s\vert_{A,B}\in\sigma,\;\s\vert_{B,C}\in\tau,\;\s\vert_{C,D}\in\upsilon\right\}
      \]
    We shall show that 
    \[
      T_{\comp{\upsilon}{(\comp{\tau}{\sigma})}} = \{\s\vert_{A,D}\suchthat\s\in\sigma\|\tau\|\upsilon\}
      \]
    The same claim will follow for $T_{\comp{(\comp{\upsilon}{\tau})}{\sigma}}$ by symmetry, completing the proof.

@@

    In any case, we have
    \begin{gather*}
      \u\vert_{A,C,D} = \s_0D_0\s_1D_1\dots = \s \\
      \u\vert_{A,B,C} = \s_0B_0\s_1B_1\dots = \t
    \end{gather*}
    It follows that $\u\in\sigma\|\tau\|\upsilon$: 
    \begin{gather*}
      \u\vert_{A,B} = \t\vert_{A,B} \in \sigma \\
      \u\vert_{B,C} = \t\vert_{B,C} \in \tau \\
      \u\vert_{C,D} = \s\vert_{C,D} \in \upsilon
    \end{gather*}
    Lastly, we have $\u\vert_{A,D} = \s\vert_{A,D} = s$, completing the proof.
  \end{proof}

@@

\begin{proposition}
  $D_{\comp\tau\sigma}$ is a well-defined divergence set.
\end{proposition}
  \begin{proof}
    $D_{\comp\tau\sigma}$ alternating by the same argument we used for $T_{\comp\tau\sigma}$, and is postfix-closed in $P_{A\implies C}^O$ by definition.  Now let $s\in D_{\comp\tau\sigma}$ and suppose that $\s\vert_{A,C}\prefix s$, where $\s\in \sigma\dv\tau$.  We need to show that there is some $O$-position $ta\prefix s$ such that $ta\in D_{\comp\tau\sigma}$ and $t\in T_{\comp\tau\sigma}$.

@@

    The case where $\s\vert_{A,B}\in T_\sigma$ and $\s\vert_{B,C}\in D_\tau$ is similar.
  \end{proof}

@@

\begin{proposition}
  Let $A,B,C$ be games, let $\sigma$ be a deterministic strategy for $A\implies B$ and let $\tau$ be a deterministic strategy for $B\implies C$.  Then $\comp\tau\sigma$ is a deterministic strategy.  Moreover, for any $s\in\comp\tau\sigma$, there is a unique minimal $\s\in\sigma\|\tau$ such that $\s\vert_{A,C}=s$.
\end{proposition}
  \begin{proof}
    The idea behind the proof is that determinism of $\sigma$ and $\tau$ mean that the play in the intermediate game $B$ is completely determined (since every move played in $B$ is either a $P$-move in $A\implies B$ or a $P$-move in $B\implies C$.  For full details, see \cite{abramskyjagadeesangames}.
  \end{proof}

@@

\begin{proposition}
  Let $A,B,C$ be games and let $\sigma\from A\implies B$ and $\tau\from B\implies C$ be compact deterministic strategies.  Then $\comp\tau\sigma$ is compact.
\end{proposition}
  \begin{proof}
    Let $s\in M_{A\implies C}^\omega$ be an infinite sequence such that $t\in\comp\tau\sigma$ for all finite prefixes $t\ppprefix s$.  For each finite prefix $t$ of $s$, there is a unique minimal sequence $\t\in\sigma\|\tau$ such that $\t\vert_{A,C}=t$.  Writing $s$ as the limit of its finite $P$-prefixes
    \[
      t_1 \prefix t_2 \prefix t_3 \prefix \dots
      \]
    we build up a sequence
    \[
      \t_1 \prefix \t_2 \prefix \t_3 \prefix \dots
      \]
    where $\t_i$ is the unique minimal sequence in $\sigma\|\tau$ such that $\t_i\vert_{A,C}=t_i$.  We have $\t_i\prefix\t_{i+1}$ because if we remove the last move from $\t_{i+1}$ then the resulting sequence $\u_i$ satisfies $\u_i\vert_{A,C}=t_i$ and therefore $\t_i\prefix\u_i\prefix\t_{i+1}$ by minimality of $\t_i$.

    Then the limit of the sequences $\t_i\vert_{A,B}$ must be contained in $T_\sigma\cup D_\sigma$, since $\sigma$ is compact, and similarly the limit of the sequences $\t_i\vert_{B,C}$ must be contained in $T_\tau\cup D_\tau$, since $\tau$ is compact.  It follows that the limit $\t$ of the sequences $\t_i$ is contained in $\sigma\|\tau\cup\sigma\dv\tau$, and therefore that $\t\vert_{A,C}=s\in T_{\comp\tau\sigma}\cup D_{\comp\tau\sigma}$.
  \end{proof}

@@

\begin{proposition}
  Let $A,B,C$ be games and let $\sigma\from A\implies B$ and $\tau\from B\implies C$ be (nondeterministic, locally compact) strategies.  Then $\comp\tau\sigma$ is locally compact.
\end{proposition}
  \begin{proof}
    Let $s\in T_{\comp\tau\sigma}$.  Then there is some sequence $\s\in\sigma\|\tau$ such that $\s\vert_{A,C}=s$.  We have $\s\vert_{A,B}\in\sigma$ and $\s\vert_{B,C}\in\tau$, so, by local compactness for $\sigma$ and $\tau$, there are compact strategies $\sigma'\from A\implies B$ and $\tau'\from B\implies C$ such that $\s\vert_{A,B}\in\sigma'$ and $\s\vert_{B,C}\in\tau'$.  After removing positions from (and possibly adding divergences to) $\sigma'$ and $\tau'$, we may assume that they are deterministic (and still compact).  Then $\comp{\tau'}{\sigma'}$ is compact and we have $\s\vert_{A,C}\in\comp{\tau'}{\sigma'}\subset\comp\tau\sigma$ and $D_{\comp\tau\sigma}\subset D_{\comp{\tau'}{\sigma'}}$.
  \end{proof}

@@

\end{proposition}
  \begin{proof}
    Suppose that $s\in\comp\tau\sigma$, that $sa\in P_{A\implies C}$ and that $s\not\in D_{\comp\tau\sigma}$.  Let $\s\in\sigma\|\tau$ be such that $\s\vert_{A,C}=s$.  By local compactness of $\sigma$ and $\tau$, there must be compact strategies $\sigma'\from A\implies B$ and $\tau'\from B\implies C$ such that $\s\vert_{A,C}\in\sigma'$ and $\s\vert_{B,C}\in\tau'$.

@@

    Since $sa$ is an $O$-position, one of $\s a\vert_{A,B}$ and $\s a\vert_{B,C}$ must be an $O$-position (by transitivity of $\Rightarrow$).  Suppose without loss of generality that $\s a\vert_{A,B}$ is an $O$-position.  Since $\s a\vert_{A,B}\not\in D_{\sigma'}$, there must be some move $b$ such that $\s a\vert_{A,B}b\in\sigma'$, since $\sigma'$ satisfies the diverge-or-reply rule.  Then we have $\s ab\vert_{A,B}\in\sigma'$.  Moreover, Lemma \ref{signProfilesLemma} tells us that $\s a\vert_{B,C}$ must be a $P$-position, so $\s ab\vert_{B,C}=\s\vert_{B,C}\in\tau'$.  Therefore, $\s ab\in\sigma'\|\tau'$.  By maximality of $\s a$, $b$ must be a move in $A$ and therefore we have $sab\in\comp{\tau'}{\sigma'}\subset\comp\tau\sigma$.
  \end{proof}

@@

\begin{proposition}
  Let $A,B,C,D$ be games, let $\sigma$ be a strategy for $A\implies B$, let $\tau$ be a strategy for $B\implies C$ and let $\upsilon$ be a strategy for $C\implies D$.  Then 
  \[
    D_{\comp\upsilon{(\comp\tau\sigma)}} = D_{\comp{(\comp\upsilon\tau)}\sigma}
    \]
\end{proposition}
  \begin{proof}
    Let
    \[
      D = \left\{\s\in \seqs{(M_A \cprd M_B \cprd M_C \cprd M_D)} \;\middle|\;\mbox{\pbox{\textwidth}{\flushRight
        either \textbf{1} $\s\vert_{A,B}\in D_\sigma$, $\s\vert_{B,C}\in T_\tau$ and $\s\vert_{C,D}\in T_\upsilon$ \\
        or \textbf{2} $\s\vert_{A,B}\in T_\sigma$, $\s\vert_{B,C}\in D_\tau$ and $\s\vert_{C,D}\in T_\upsilon$ \\
        or \textbf{3} $\s\vert_{A,B}\in T_\sigma$, $\s\vert_{B,C}\in T_\tau$ and $\s\vert_{C,D}\in D_\upsilon$
      }}\right\}
      \]
    We claim that $D_{\comp\upsilon{(\comp\tau\sigma)}}=\pocl_{A\implies D}^O\left(\{\s\vert_{A,D}\suchthat\s\in D\}\right)$.  The same claim will hold for $D_{\comp{(\comp\upsilon\tau)}\sigma}$ by symmetry, completing the proof.

@@

    By the argument from Proposition \ref{traceAssociativity}, we may interleave these two sequences $\t$ and $u$ to form a sequence $\s\in\seqs{(M_A\cprd M_B\cprd M_C\cprd M_D)}$ such that $\s\vert_{A,B,C}=\u$ and $\s\vert_{A,C,D}=\t$.  Then it is clear that $\s\in D$ in all cases and we have $\s\vert_{A,D}=\t\vert_{A,D}\prefix s$.
  \end{proof}

@@

\begin{theorem}
  There is a category $\G$ where the objects are games as defined above, morphisms from a game $A$ to a game $B$ are given by nondeterministic strategies $(T_\sigma, D_\sigma)$ for $A\implies B$ and the composition and identity morphisms are as defined above.
\end{theorem}
  \begin{proof}
    All that is left to do is to show that $\id_A$ really is an identity morphism.  We shall show that if $A$ and $B$ are games and $\sigma$ is a strategy for $A\implies B$ then $\comp\sigma{\id_A}=\sigma$; it will follow by a symmetrical argument that $\comp{\id_B}\sigma = \sigma$.

@@

    Conversely, suppose that $s\in D_\sigma$.  We know already that there is some $\s\in\id_A\|\sigma$ such that $\s\vert_{A_1,A_2}\in\id_A$ and $\s\vert_{A_2,B}\in\sigma$.  Then $\s\vert_{A,C}\in D_{\comp\sigma{\id_A}}$ by definition.
  \end{proof}

@@

We make the category $\G$ into a symmetric monoidal closed category with tensor product given by $\tensor$ and linear implication given by $\implies$.  At the level of morphisms, the action of $\blank\tensor\blank$ is defined as follows: if $A,B,C,D$ are games, $\sigma$ is a strategy for $A\implies C$ and $\tau$ is a strategy for $B\implies D$, then $\sigma\tensor\tau$ is the strategy for $(A\tensor B)\implies (C\tensor D)$ given by playing the strategy $\sigma$ in $A$ and $C$ and playing the strategy $\tau$ in $B$ and $D$:
\[
  \sigma\tensor\tau = \left\{s\in P_{(A\tensor B)\implies (C\tensor D)}\suchthat s\vert_{A,C}\in\sigma,\;s\vert_{B,C}\in\tau\right\}
  \]
The divergences for $\sigma\tensor\tau$ are defined using the divergences for $\sigma$ and $\tau$:
\[
  D_{\sigma\tensor\tau} = \pocl^O\left(\left\{s\in P_{(A\tensor B)\implies (C\tensor D)}\;\middle|\;\mbox{\pbox{\textwidth}{\flushRight
    either $s\vert_{A,C}\in D_\sigma$ and $s\vert_{B,D}\in T_\tau$ \\
    or $s\vert_{A,C}\in T_\sigma$ and $s\vert_{B,D}\in D_\tau$
    }}\right\}\right)
  \]
\begin{proposition}
  $\sigma\tensor\tau$ is a valid strategy for $(A\tensor B)\implies (C\tensor D)$
\end{proposition}
  \begin{proof}
    $\sigma\tensor\tau$ is non-empty and prefix-closed, since $\sigma$ and $\tau$ are.

@@

    We can show that
    \[
      D_{\comp{(\sigma\tensor\tau)}{(\sigma'\tensor\tau')}} = D_{(\comp\sigma{\sigma'})\tensor(\comp\tau{\tau'})}
      \]
    using a similar argument.
  \end{proof}

@@

\begin{proposition}
  $\oc A$ is a well formed game.
\end{proposition}
  \begin{proof}
    Exactly the same argument as in Proposition \ref{TensorImpliesWellFormed}.  We need to prove a version of Lemma \ref{signProfilesLemma} which will tell us that if $s\in P_{\oc A}$ then either $\zeta_A(s\vert_n)=O$ for at most one value of $n$.  
  \end{proof}

@@

\subsection{Consistency}
